---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Set up working directory
```{r, echo=TRUE}
setwd("~/R training/Repro Res/RepData_PeerAssessment1")
dir() # check the files in working dir
```
## load standard libraries
```{r, echo=TRUE}
## load libraries----

library(dplyr)
library(data.table)
library(ggplot2)
library(lubridate)
library(tidyr)

```

## Loading and preprocessing the data

```{r, echo=TRUE}
activity <- read.csv("./activity/activity.csv")
str(activity)
head(activity)
```
### Calculate total steps for each day

```{r, echo=TRUE}
histo <- activity %>%
                group_by(date) %>%
                        summarize(total_o = sum(steps, na.rm = TRUE)) 
# total_o represents total steps from original datasets
histo
```

## What is mean total number of steps taken per day?
### Plot Histogram for the total steps taken each day
```{r, echo = TRUE}
histo %>% ggplot(aes(x=total_o)) +
geom_histogram(binwidth=600, fill="blue", color="blue", alpha=0.9) +
        ggtitle("Bin size = 3") +
        labs(title = "Histogram of Total Steps Each Day", x = "Total Steps", y = "Frequency") +
        theme_minimal() +
        theme(plot.title = (element_text(size=15)
        ))
histo
```
### Calculate and report the mean and median of the total number of steps taken per day

```{r, echo=TRUE}
mean_median_original <- histo %>%
        summarize(
                mean_steps_pd = mean(total_o, na.rm = TRUE),
                median_steps_pd = median(total_o, na.rm = TRUE)
                )
mean_median_original
```

## What is the average daily activity pattern?
### Calculate the average steps in each 5 min time interval over Oct and Nov 2012
```{r, echo=TRUE}
avg_steps <- activity %>%
                group_by(interval) %>%
                summarize(mean = mean(steps, na.rm = TRUE))
avg_steps
```
### Create a time series plot (decided to used ggplot) avg_steps (y-axis and interval (x-axis)
```{r, echo=TRUE}
p <- ggplot(avg_steps, aes(x = interval, y = mean)) +
  geom_line(color = "steelblue", linewidth = 1) +   # Line (type = 1)
  labs(title = "Average Daily Activity Pattern", x = "Interval", y = "Average Steps per Interval") +
  theme_minimal()
p

```

### Which 5-minute interval contain maximum average steps over Oct and Nov 2012
```{r, echo=TRUE}
summary(avg_steps$mean) ## to see max value in summary stats
max_steps_interval <- avg_steps[which.max(avg_steps$mean), "interval"]
max_steps_interval

max_steps <- avg_steps %>% # to confirm the value and corresponding interval in the same file
        filter(interval==835)
max_steps

```
## Imputing missing values
### total number of days/intervals where missing values is NA
```{r, echo=TRUE}
rows_NA <- sum(!complete.cases(activity))  # rows with NA
rows_NA

```
### Impute NA with average for the interval in activity dataset
```{r, echo = TRUE}
# Add interval-wise mean to activity dataset by interval
activity_impute <- activity %>%
  left_join(avg_steps, by = "interval") %>%
        mutate(
    steps_imputed = coalesce(steps, mean)  
    # replaces NA with mean value by interval
  )
head(activity_impute)
```
### Repeat Histogram and mean median calculated with inputed NA dataset
### Calculate total steps for each day

```{r, echo=TRUE}
histo_i <- activity_impute %>%
                group_by(date) %>%
                        summarize(total_i = sum(steps_imputed, na.rm = TRUE)) 
# total_i refers to total steps calculated using imputed dataset
histo_i
```

## What is mean total number of steps taken per day?
### Plot Histogram for the total steps taken each day
```{r, echo = TRUE}
histo_i %>% ggplot(aes(x=total_i)) +
geom_histogram(binwidth=600, fill="blue", color="blue", alpha=0.9) +
        ggtitle("Bin size = 3") +
        labs(title = "Histogram of Total Steps Each Day - Imputed Dataset", x = "Total Steps", y = "Frequency") +
        theme_minimal() +
        theme(plot.title = (element_text(size=15)
        ))
histo_i
```
### Calculate and report the mean and median of the total number of steps taken
using imputed dataset

```{r, echo=TRUE}
mean_median_imputed <- histo_i %>%
        summarize(
                mean_steps_pd = mean(total_i, na.rm = TRUE),
                median_steps_pd = median(total_i, na.rm = TRUE)
                )
mean_median_imputed
```

### Compare Mean Mediantotal daily steps - Original vs Imputed Datasets
```{r, echo=TRUE}
compare <- bind_rows(mean_median_original, mean_median_imputed)
compare
# Major differences in mean and minor differences in median value were observed
```
## Are there differences in activity patterns between weekdays and weekends?
### Create a factor variable for "weekday" or "weekend"
```{r, echo=TRUE}
activity2 <- activity_impute %>%
  mutate(date = as.Date(date)) %>%
  mutate(
    day_type = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
  )
head(activity2)
```

### Calculate the average number of steps across all weekday days or weekend days
```{r, echo=TRUE}
steps_day <- activity2 %>%
                group_by(day_type, interval) %>%
                summarize(mean = mean(steps_imputed, na.rm = TRUE))
head(steps_day)
summary(steps_day$mean)
```
### Create a panel plot to compare activity weekday vs weekend
# Note: I used ggplot instead of base or lattice plotting system, hence looks slightly different
# from the example showed in the assignment
```{r, eco=TRUE}
ggplot(steps_day, aes(x = interval, y = mean, group = day_type, color = day_type)) +
  geom_line() +
  facet_wrap(~ day_type, ncol = 1) + 
  theme_minimal() +
  labs(title = "Activity Patterns Over Weekdays vs Weekends", x = "Interval", y = "Average Number of Steps")
```